A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-20 linear.tex
fixed SI results. need to use largrangian pert properly

\section{Compressible streaming instability}\label{si}
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

{\bf note: kappa2 accounts for dust effect assuming smallh=0.05}
We solve Eq. \ref{streaming_dispersion} for selected cases where the  
growth rates of the streaming instability have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_x = \eta r k_x$ and similarly for $K_z$, where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as 
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

Table \ref{si_compare} compares the eigenfrequencies obtained from the
one-fluid dispersion relation to that from the above studies. As  
expected eigenfrequencies agree better with decreasing $\tstop$ since
in that limit the mixture behaves more like a single fluid. Most 
importantly, we find the work done $\mathcal{W}>0$ in all cases, and
hence find growing oscillations. This implies the streaming instability stems
from the gas pressure evolution lagging behind that of the total (and hence dust) density.

Notice also eigenfrequencies agree better for 
smaller pressure-density phase lags. While this is generally true with decreasing $\tstop$, other 
problem parameters may affect the magnitude of this phase lag, and hence the validity of the one-fluid model.  
For example, `linA' has a larger $\tstop$ than `linC', but linA's one and 
two-fluid growth rates differ by $\sim 1\%$; compare to $\sim 10\%$ error for linC. 
However, linC has a larger phase lag. This means that the effective coupling strength between dust and gas 
is problem-dependent.

@@

\begin{deluxetable*}{llrrrrrr}
  \tablecolumns{8}
  \tablecaption{Selected eigenfrequencies for the streaming
    instability. \label{si_compare}
  }
  \tablehead{
    \colhead{Mode} & 
    \colhead{$\tau_\mathrm{s}\OmK$} &
    \colhead{$\epsilon$} &
     \colhead{$K_{x,z}$} &
      \colhead{$\sigma/\OmK$ (two-fluid)} &
    \colhead{$\sigma/\OmK$ (one-fluid)} &
    \colhead{$\mathcal{W}$ (arbitrary units)} &
      \colhead{Pressure lag}
  }
\startdata
 linA, \cite{youdin07b} &  $0.1$       & 3.0 & 30    & $-0.3480 +
 0.4190\ii$ & $0.3640 + 0.4249\ii$ & $0.026$ & $42\degr$\\

linB, \cite{youdin07b} & $0.1$        &  0.2 & 6 & $0.4999 +
0.0155\ii$&   $-0.4981 + 0.0054\ii$  & $1.75$ & $88\degr$ \\

linC,  \cite{bai10b}  & $10^{-2}$   &  2.0 & 1500&   $0.1049 +
0.5981\ii$   &  $0.1338 + 0.6650\ii$  & $3.57\times10^{-4}$& $78\degr$ \\

linD, \cite{bai10b} &  $10^{-3}$   &  2.0 & 2000 & $0.3225 + 
0.3154\ii$& $0.3219 + 0.3154\ii$ &  $1.85\times10^{-4}$ & $21\degr$ 
\enddata
\end{deluxetable*}

>>>>

\section{Compressible streaming instability}\label{si}
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

{\bf note: kappa2 accounts for dust effect assuming smallh=0.05}
We solve Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rate have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_x = \eta r k_x$ and similarly for $K_z$, where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as 
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

Table \ref{si_compare} compares the eigenfrequencies obtained from the
one-fluid dispersion relation to that from the above studies. As  
expected eigenfrequencies agree better with decreasing $\tstop$ since
in that limit the mixture behaves more like a single fluid. Most 
importantly, we find the work done $\mathcal{W}>0$ in all cases, and
hence find growing oscillations.

We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\mathcal{I}^2}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. Interestingly, we find that for cases with
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta P \to
\Delta P - \dd P$, i.e. the advection part of the Lagragian
pressure perturbation is dominant. Conversely, for `linB' with $\epsilon
< 1$, one can take $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. This
suggests that for SI in gas-dominated disks the distinction between
Eulerian and Lagragian presure perturbations is unimportant.

{\bf note: important to use largrangian pressure pert}
In Table \ref{si_compare} we also calculate the phase difference
between the Lagragian pressure perturbation and density perturbations
as    
\begin{align*} 
\varphi = -\sgn(\omega)\arg\left(\Delta P\Delta\rho^*\right), 
\end{align*}
so that $\varphi > 0 $ indicates pressure lagging behind density, which
is true for all the cases. In particular, `linB' indicates that growth
rates vanish as $\varphi \to 0$.  
Thus the streaming instability stems  
from the gas pressure evolution lagging behind that of the total
density (and hence behind the dust density for isothermal gas).

@@

\begin{deluxetable*}{llrrrrrr}
  \tablecolumns{8}
  \tablecaption{Selected eigenfrequencies for the streaming
    instability. \label{si_compare}
  }
  \tablehead{
    \colhead{Mode} & 
    \colhead{$\tau_\mathrm{s}\OmK$} &
    \colhead{$\epsilon$} &
     \colhead{$K_{x,z}$} &
      \colhead{$\sigma/\OmK$ (two-fluid)} &
    \colhead{$\sigma/\OmK$ (one-fluid)} &
    \colhead{$\mathcal{W}$ (arbitrary units)} &
      \colhead{$\Delta P$ lag}
  }
\startdata
 linA, \cite{youdin07b} &  $0.1$       & 3.0 & 30    & $-0.3480 +
 0.4190\ii$ & $0.3640 + 0.4249\ii$ & $0.90$ & $30\degr$\\

linB, \cite{youdin07b} & $0.1$        &  0.2 & 6 & $0.4999 +
0.0155\ii$&   $-0.4981 + 0.0054\ii$  & $1.54$ & $1.2\degr$ \\

linC,  \cite{bai10b}  & $10^{-2}$   &  2.0 & 1500&   $0.1049 +
0.5981\ii$   &  $0.1338 + 0.6650\ii$  & $0.15$& $11\degr$ \\

linD, \cite{bai10b} &  $10^{-3}$   &  2.0 & 2000 & $0.3225 + 
0.3154\ii$& $0.3219 + 0.3154\ii$ &  $1.28$ & $22\degr$ 
\enddata
\end{deluxetable*}
